Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1) \note{\scriptsize Authors, please add/check middle initials and affiliations}

### 2014-04-21 WSSSPE1_report/WSSSPE1_summary_report.tex
Rework of cross-cutting issue sust. def.

Two issues were raised in a number of the workshops sessions, defining sustainability and career paths.  This section brings those discussions together.

@@

%\note{So far this reads a lot like a collection of thoughts only, without ``red line''.}
%\note{Dan: assuming the previous note is from FL, what do you propose?}
The question of what was meant by ``sustainability'' in the context of software
came up in many different parts of the workshop, specifically in the first
keynote (\S\ref{sec:keynote1}), the Developing and Maintaining Software panel,
and the Policy panel.

In the opening keynote, Philip Bourne suggested that perhaps sustainability can
be defined as the effort that happens to make the essential things continue.
This leads to having to decide what it is that we want to sustain, whether what
we want to sustain is valuable, and finally, who would care if it went away,
and how one measures how much they care.

In the Developing and Maintaining Software panel, there was some discussion on
this topic: what does sustainability mean? It was pointed out that OSS Watch
proposes a Software Sustainability Maturity Model to address the issue of how
sustainability a particular element of software is, and says that this
sustainability is important. ``When choosing software for procurement or
development reuse - regardless of the license and development model you will
use - you need to consider the future. While a software product may satisfy
today's needs, will it satisfy tomorrow's needs? Will the supplier still be
around in five years' time? Will the supplier still care for all its customers
in five years' time? Will the supplier be responsive to bug reports and feature
requests? In other words, is the software sustainable?''~\cite{OSS-ssmm-web}

Attendees suggested that a key question that the definition of sustainability
is one issue on which the community needs to agree, and that ideally, an
initial definition would be determined during the workshop, or at least some
progress would be made towards this goal.  Another topic that came up is what
the goal of sustainability is.  Perhaps it is reproducible science, or
persistence, or quality, or something else.  Similarly, some attendees want to
understand how success in sustainability is measured.  How does a group of
developers know when they have actually achieved sustainable software? This led
to a comment that sustainability should be addressed throughout the full
software life cycle.

Another topic that came up during the Developing and Maintaining Software panel
is the relationship of sustainability to other software attributes.  Attendees
asked ``what is the relationship between sustainability and provenance?'' And,
``is usability separate from sustainability or a fundamental part of it?''

In addition, the Policy panel had a large amount of discussion about defining
sustainability, as one of the subtopics in that panel was Modeling
Sustainability, and modeling requires defining what will be modeled.

In this subtopic, two papers included discussion of the definition of
sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned that this is a
rather ambiguous concept, and that the lack of an accepted definition gets in
the way of integrating the concept into software engineering. They suggest that
sustainability is a non-functional requirement, and that the quality of
software architectures determines sustainability. They then propose that
sustainability is a consequence of a set of central quality attributes:
extensibility, interoperability, maintainability, portability, reusability, and
scalability. Finally, they develop an architecture evaluation framework based
on scenarios that help to illuminate how to measure quality or sustainability
at the architectural level.

Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could be
used to measure software sustainability, and though these questions might
falsely lead to yes or no answers, the complete set would determine a range of
values for sustainability. These questions are:
\begin{itemize}
\item Will the software continue to provide the same functionality in the future,
      even if the environment (other parts of the infrastructure) changes?
\item Is the functionality and usability clearly explained to new users?
\item Do users have a mechanism to ask questions and to learn about the software?
\item Will the functionality be correct in the future, even if the environment changes?
\item Does it provide the functionality that existing and future users want?
\item Does it incorporate new science/theory/tools as they develop?
\end{itemize}

Lenhardt~\cite{lenhardt-wssspe1-panel} summarized the contributions of the
Modeling Sustainability papers in the panel. As shown in
Table~\ref{tab:defining-sustainability}, for each paper he discussed what
software meant, whether there was a definition of sustainability, and what the
approach was to either understand or evaluate sustainability.

\begin{table}[t]
  \begin{scriptsize}
    \begin{center}
      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from~\cite{lenhardt-wssspe1-panel}.}
      \label{tab:defining-sustainability}
      \begin{tabular}{|p{2.3cm}|p{3.6cm}|p{4.4cm}|p{4.8cm}|}
                \hline
{\bf Paper/Authors}
& {\bf Software}
& {\bf Sustainability}
& {\bf Approach to Understand or Evaluate Sustainability} \\
                \hline
Calero, et al.~\cite{Calero_WSSSPE}
& General notion of software. Not explicitly defined.
& Sustainability is linked to quality.
& Add to ISO \\
                \hline
Venters, et al.~\cite{Venters_WSSSPE}
& Software as science software; increasingly complex; service-oriented computing
& Extensibility, interoperability, maintainability, portability, reusability, scalability, efficiency
& Use various architecture evaluation approaches to assess sustainability \\
                \hline
Pierce, et al.~\cite{Pierce_WSSSPE}
& Cyberinfrastructure software
& Sustainable to the extent to which there is a community to support it
& Open community governance \\
                \hline
Katz and Proctor~\cite{Katz_WSSSPE}
& E-research infrastructures (i.e. cyberinfrastructure)
& Persisting over time, meeting original needs and projected needs
& Equates models for the creation of software with sustaining software \\
                \hline
Lenhardt, et al.~\cite{Lenhardt_WSSSPE}
& Broadly defined as software supporting science
& Re-use; reproducible science
& Comparing data management life cycle to software development life cycle \\
                \hline
Weber, et al.~\cite{Weber_WSSSPE}
& Software broadly defined; a software ecosystem
& Software niches
& Ecological analysis and ecosystem \\
                \hline
     \end{tabular}
    \end{center}
  \end{scriptsize}
\end{table}

>>>>

Two issues, how to define software sustainability, and how career
paths (or their lack) interact with achieving it, were raised across
the workshop's panels.  This section aims to synthesize these
discussions from different perspectives.

@@

What is, or should be meant by ``sustainability'' in the context of
software came up in many different parts of the workshop, specifically
in the first keynote (\S\ref{sec:keynote1}), the Developing and
Maintaining Software panel (\S\ref{sec:devel}), and the Policy panel
(\S\ref{sec:policy}). It quickly became clear that at present there is
no consensus among the community, whether within or accross
disciplines, on what this definition should be, and that a variety of
different defintiions were being assumed, used, or sometime expressly
articulated by contributors and attendees. However, some concepts,
particularly relating sustainability to change over time, were also
evidently held in common. This common notion is, for example, captured
in the definition used by the UK's Scientific Software Sustainability
Institute, quoted in~\cite{Venters_WSSSPE}: ``software you use today
will be available---and continue to be improved and supported---in the
future''. Pierce et al~\cite{Pierce_WSSSPE} express this idea as
software that continues to serve its users.

Philip Bourne, too, used the relation to change over time when he
suggested in his opening keynote that sustainability can perhaps be
defined as the effort needed to make the essential things continue.
This leads to having to decide what it is that we want to sustain,
whether what we want to sustain is valuable, and finally, who would
care and how much if it went away. As was pointed out during a
discussion session, OSS Watch, an advisory service for issues relating
to free and open source software in the UK, proposes a Software
Sustainability Maturity Model to address the issue of what level of
sustainability a particular element of software needs to have, and
where this is important. It, too, expresses sustainability in relation
to change over time:
\begin{quote}``When choosing software for procurement or
development reuse --- regardless of the license and development model you will
use --- you need to consider the future. While a software product may satisfy
today's needs, will it satisfy tomorrow's needs? Will the supplier still be
around in five years' time? Will the supplier still care for all its customers
in five years' time? Will the supplier be responsive to bug reports and feature
requests? In other words, is the software sustainable?''~\cite{OSS-ssmm-web}
\end{quote}

Attendees suggested that having a definition of sustainability on
which the community can agree is key.  A related question that was
raised is what the goal of sustainability should be, with a wide range
of possible answers, including more reproducible science, software
persistence, and quality.  And given a goal of sustainability, how
would success in achieving it be measured?  How would one know that a
piece of software has reached sustainability? As a commenter
suggested, this may mean that sustainability should be addressed
throughout the full software life cycle.  It was also pointed out that
software sustainability is not isolated from other attributes of
scientific software and its use, such as usability, and provenance.

Because part of the Policy panel (\S\ref{sec:policy}) was about
modeling sustainability, and modeling requires defining what will be
modeled, this panel saw particular attention to the questions
surrounding the definition of sustainability. Two papers, Venters et
al.~\cite{Venters_WSSSPE} and Katz and Proctor~\cite{Katz_WSSSPE},
specifically discuss the issue.

According to Venters et al.~\cite{Venters_WSSSPE}, sustainability is a
rather ambiguous concept, and the lack of an accepted definition
hampers integrating the concept into software engineering. They
suggest that sustainability falls under the category of non-functional
requirements, and that a software's sustainability is a consequence of
a set of central software architecture quality attributes, including
extensibility, interoperability, maintainability, portability,
reusability, and scalability. They also propose an evaluation
framework with which quality and sustainability could be measured at
the architectural level.

Katz and Proctor~\cite{Katz_WSSSPE} propose a set of questions that could be
used to measure software sustainability:
\begin{itemize}
\item Will the software continue to provide the same functionality in the future,
      even if the environment (other parts of the infrastructure) changes?
\item Is the functionality and usability clearly explained to new users?
\item Do users have a mechanism to ask questions and to learn about the software?
\item Will the functionality be correct in the future, even if the environment changes?
\item Does it provide the functionality that existing and future users want?
\item Will it incorporate new science, theory, or tools as they develop?
\end{itemize}
Despite their phrasing, these questions are not intended to be given
simplistic yes or no answers, and it is the complete set rather than
any individual one that would determine where in the range of
sustainability a software falls.

Table~\ref{tab:defining-sustainability} reproduces a summary created
by C.~Lenhardt ~\cite{lenhardt-wssspe1-panel}, which shows for each contribution to the
Modeling Sustainability panel what it meant when referring to
software, how it defined software sustainability, and which
approach it suggested to understand or evaluate sustainability.
\begin{table}[t]
  \begin{scriptsize}
    \begin{center}
      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from~\cite{lenhardt-wssspe1-panel}.}
      \label{tab:defining-sustainability}
      \begin{tabular}{|p{2.3cm}|p{3.6cm}|p{4.4cm}|p{4.8cm}|}
                \hline
{\bf Paper/Authors}
& {\bf Software}
& {\bf Sustainability}
& {\bf Approach to Understand or Evaluate Sustainability} \\
                \hline
Calero, et al.~\cite{Calero_WSSSPE}
& General notion of software. Not explicitly defined.
& Sustainability is linked to quality.
& Add to ISO \\
                \hline
Venters, et al.~\cite{Venters_WSSSPE}
& Software as science software; increasingly complex; service-oriented computing
& Extensibility, interoperability, maintainability, portability, reusability, scalability, efficiency
& Use various architecture evaluation approaches to assess sustainability \\
                \hline
Pierce, et al.~\cite{Pierce_WSSSPE}
& Cyberinfrastructure software
& Sustainable to the extent to which there is a community to support it
& Open community governance \\
                \hline
Katz and Proctor~\cite{Katz_WSSSPE}
& E-research infrastructures (i.e. cyberinfrastructure)
& Persisting over time, meeting original needs and projected needs
& Equates models for the creation of software with sustaining software \\
                \hline
Lenhardt, et al.~\cite{Lenhardt_WSSSPE}
& Broadly defined as software supporting science
& Re-use; reproducible science
& Comparing data management life cycle to software development life cycle \\
                \hline
Weber, et al.~\cite{Weber_WSSSPE}
& Software broadly defined; a software ecosystem
& Software niches
& Ecological analysis and ecosystem \\
                \hline
     \end{tabular}
    \end{center}
  \end{scriptsize}
\end{table}
